A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-04-05 analysis/paper/paper.Rmd
update paper

In this study, we evaluated a total of  `r nrow(filter(methods, evaluated, type == "algorithm"))` methods in terms of cellular ordering, trajectory topology, code quality and user friendliness. Our results indicate that there is no "one-size-fits-all" method, but that the choice of method mostly depends on the topology present in the data. We therefore developed a set of guidelines to aid users select the most appropriate method, based on their prior knowledge of the trajectory present in the data. While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases. We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extendable and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).

@@

<!-- §3: Problem statement -->
Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness. Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation of TI methods, which compares a large set of methods on a large set of datasets, is still lacking. This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about which method would optimally solve their problem. Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.

<!-- §4: Our study -->
In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`). The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`). In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories. The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation. Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("ti_evaluation_overview_v3.svg", experiment_id="manual_figures"),
  ref_id = "ti_evaluation_overview", 
  caption_main = "Overview of several key aspects of the evaluation.", 
  caption_text = "a) A schematic overview of our evaluation pipeline. b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods. c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
)
```

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. We then defined a set of metrics for comparing the similarities of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "ti_evaluation_overview", "c")`). The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). In order to generate synthetic datasets, we simulated gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.

@@

<!-- Errors-->
Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g. Monocle DDRTree, cellTree Gibbs, SLICE), meaning these methods could rank even higher if not for the failed executions. Methods that do not scale well with respect to the number of cells or genes exceeded the pre-defined time or memory limits for the largest datasets (reCAT, SCOUP, StemID, cellTree Gibbs). For a few methods, the time or memory limits were exceeded too often, making the benchmarking results difficult to compare to those of other methods (Ouija, Pseudogp).

@@

Despite their similar overall performance, the topologies predicted by the top methods differed considerably. Trajectories detected by Slingshot and cellTree tended to contain less branches, while those detected by TSCAN and AGA Pseudotime gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`). AGA Pseudotime for example frequently predicted a tree topology, even when a non-branching topology was present in the data. Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree most frequently predicted the correct tree topology (`r ref("stable", "topology_sensitivity")`). Although the top methods in the ranking were capable at discriminating linear from bifurcating trajectories (with TSCAN predicting the correct topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'tscan', trajectory_type == 'directed_linear')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'tscan', trajectory_type == 'bifurcation')$perc_perfect)` of datasets), the accuracy when more complex tree trajectories were present in the data was still very low, with the best methods, Monocle DDRTree and SCUBA, predicting the correct tree topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'mnclddr', trajectory_type == 'rooted_tree')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'scuba', trajectory_type == 'rooted_tree')$perc_perfect)` of the cases  (`r ref("stable", "topology_sensitivity")`). Inferring the correct topology without parameter tuning is therefore still an open challenge. When the data contains a complex trajectory structure, TI will currently require a considerable guidance by the user, either to optimise the parameters or to choose the method whose output best fits the user's expert knowledge.

@@

<!-- Additional factors which could influence performance -->
Apart from the trajectory types, we also investigated several other factors which could influence the performance of a method. Although we provided prior information to those methods that require it, we did not observe any major difference in method performance between these methods and those that did not receive any prior information (`r ref("fig", "results_overview", "a-b")`). When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance, while those which included tSNE had a lower performance ( `r ref("stable", "component_importance")`), although these tendencies were not significant.

@@

<!-- FOR METHOD DEVELOPERS -->
<!-- New tools: challenges and possibilities -->
Our study indicates that the field of trajectory inference is maturing, primarily for linear and bifurcating trajectories. Nonetheless, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable and fast tool for analysing single-cell -omics datasets with complex trajectories. Foremost, new methods should focus on improving the inference of the topology for tree and more complex topologies, and should include better ordering methods, perhaps borrowed from the top linear TI methods. Furthermore, new tools should incorporate higher standards for code assurance and documentation, to improve the friendliness towards both users and developers. To support the development of these new tools, we provide all datasets from this study [@cannoodt_goldstandardsinglecell_] and provide the code to compare methods in a series of R packages available at [https://github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).

@@

* **Random:** A random network with 50 milestones according to the Barabási–Albert model is generated, and cells are placed randomly along the network. This method provides a baseline performance to compare against; each other TI method should be able to obtain a higher benchmark score than this control method.
* **Component 1:**: This method returns the first component of a PCA dimensionality reduction as a linear trajectory. This method was included as some researchers indeed use this approach to obtain a pseudotime ordering of the cells.
* **Angle:** Similar to the previous method, this method computes the angle with respect to the origin in a two-dimensional PCA plot, and returns a cyclical trajectory. As reCAT was the only method specifically designed to produce cyclical trajectories, we believed it would be important to add this simple approach.

## Real datasets
We gathered `r sum(tasks_info$task_source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`). The scripts to download and process these datasets will be made available on our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalised and filtered using a common pipeline, discussed later. We determined a reference standard for every dataset using labelling provided by the author's, and classified the standards into gold and silver based on whether this labelling was determined by the expert using the expression (silver standard) or using other external information (such as FACS or the origin of the sample, gold standard) (`r ref("stable", "datasets")`).

@@

```{r}
add_sfig(
  figure_file("edge_flip_comparison.svg", experiment_id="3-score_rules/4-edge_flip"), 
  "edge_flip_comparison", 
  paste0("Edge flip score values for common trajectory topologies"), 
  15, 
  5
)
```

@@

```{r}
qc_references <- readRDS(result_file("checks.rds", "4-method_characterisation")) %>% 
  .$references %>% 
  str_split("[, \n]+") %>% 
  unlist() %>% 
  unique() %>% 
  discard(is.na) %>%
  glue::collapse(., ";") %>%
  paste0("[", ., "]")
```

We created a transparent scoring scheme to check the quality of each method (`r ref("stable", "qc_checks")`), based on several existing tool quality and programming guidelines in literature and online `r qc_references` . The goal of this quality control in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods. The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items. The availability category checks whether the method is easily available, whether the code and dependencies can be easily installed, and how the method can be used. The code quality assesses the quality of the code both from a user perspective (function naming, dummy proofing and availability of plotting functions) and a developer perspective (consistent style and code duplication). The code assurance category is frequently overlooked, and checks for code testing, continuous integration [@beaulieu-jones_reproducibility_2017] and an active support system. The documentation category checks the quality of the documentation, both externally (tutorials and function documentation) and internally (inline documentation). The behaviour category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output. Finally, we also assessed certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of dataset in which the usefulness of the method was shown, and the scope of method evaluation in the paper.

@@

## Trajectory types
We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria. These trajectory types start from the most general trajectory type, a disconnected directed graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cyclical. A disconnected directed graph is a graph in which only one edge can exist between two nodes. A directed graph is a disconnected graph in which all nodes are (indirectly) connected. A directed acyclic graph is a graph containing no cycles. A rooted tree is a directed acyclic graph containing no convergences (no nodes with in-degree higher than 1). A rooted binary tree is a rooted tree with every node having maximally two outgoing edges. A convergence is a directed acyclic graph in which only one node has a degree larger than one and this same node has an indegree of one. A multifurcation is a rooted tree in which only one node has a degree larger than one. A bifurcation is a multifurcation in which only one node has a degree equal to 3. A directed linear graph is a graph in which no node has a degree larger than 3. Finally, a directed cycle is a directed graph in which no node has a degree larger than 3 but contains one cycle. For every directed trajectory type, a corresponding undirected trajectory type can also be defined, with directed acyclic graphs and directed graphs merging to undirected graph, and convergence and bifurcation merging to a simple fork. In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with the only exception being DPT (limited to bifurcating trajectories).

## Data and code availability
The datasets used in this study are deposited in zenodo [doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533). Code is available as several R packages on github [github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).

```{r "supplementary", echo=FALSE, results='asis'}
if (params$table_format == "html") 
  source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
write_rds(lst(sfigs, stables), derived_file("sfigs_stables.rds", "paper"))
```

>>>>

In this study, we evaluated a total of  `r nrow(filter(methods, evaluated, type == "algorithm"))` methods in terms of cellular ordering, trajectory topology, code quality and user friendliness. Our results indicate that there is no "one-size-fits-all" method, but that the choice of method mostly depends on the topology present in the data. We therefore developed a set of guidelines to aid users select the most appropriate method, based on their prior knowledge of the trajectory present in the data. While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases. We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).

@@

<!-- §3: Problem statement -->
Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness. Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation of TI methods, which compares a large set of methods on a large set of datasets, is still lacking. This is problematic, as new users to the field are confronted with an overwhelming choice of TI methods, without a clear idea about which method would optimally solve their problem. Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.

<!-- §4: Our study -->
In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`). The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`). In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories. The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation. Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("overview_evaluation_v4.svg", experiment_id="manual_figures"),
  ref_id = "ti_evaluation_overview", 
  caption_main = "Overview of several key aspects of the evaluation.", 
  caption_text = "a) A schematic overview of our evaluation pipeline. b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods. c) Trajectories are automatically classified into one of nine trajectory types. d) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
)
```

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`). We defined a set of metrics for comparing the similarities of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "ti_evaluation_overview", "d")`). The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). In order to generate synthetic datasets, we simulated gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.

@@

<!-- Errors-->
Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g. Monocle DDRTree, cellTree Gibbs, SLICE), meaning these methods could rank even higher if not for the failed executions. Other methods (reCAT, SCOUP, StemID, cellTree Gibbs) exceeded the pre-defined time or memory limits when applied to the largest datasets, and therefore were not scalable to large numbers of cells or genes. For a few methods, the time or memory limits were exceeded too often, making the benchmarking results difficult to compare to those of other methods (Ouija, Pseudogp).

@@

Despite their similar overall performance, the topologies predicted by the top methods differed considerably. Trajectories detected by Slingshot and cellTree tended to contain less branches, while those detected by TSCAN and AGA Pseudotime gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`). AGA Pseudotime for example frequently predicted a tree topology, even when a non-branching topology was present in the data. Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree most frequently predicted the correct tree topology (`r ref("stable", "topology_sensitivity")`). Although the top methods in the ranking were capable at discriminating linear from bifurcating trajectories (with TSCAN predicting the correct topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'tscan', trajectory_type == 'directed_linear')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'tscan', trajectory_type == 'bifurcation')$perc_perfect)` of datasets), the accuracy when more complex tree trajectories were present in the data was still very low, with the best methods, Monocle DDRTree and SCUBA, predicting the correct tree topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'mnclddr', trajectory_type == 'rooted_tree')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'scuba', trajectory_type == 'rooted_tree')$perc_perfect)` of the cases  (`r ref("stable", "topology_sensitivity")`). Inferring the correct topology without parameter tuning is therefore still an open challenge. Currently, inferring a complex trajectory structure still requires a considerable user input, either by setting optimal parameters, or by subjectively choosing a method whose results best fit the user’s expert knowledge.

@@

<!-- Additional factors which could influence performance -->
Apart from the trajectory types, we also investigated several other factors that could influence the performance of a method. Although we provided prior information to those methods that require it, we did not observe any major difference in method performance between these methods and those that did not receive any prior information (`r ref("fig", "results_overview", "a-b")`). When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance, while those which included tSNE had a lower performance ( `r ref("stable", "component_importance")`), although these tendencies were not significant.

@@

<!-- FOR METHOD DEVELOPERS -->
<!-- New tools: challenges and possibilities -->
Our study indicates that the field of trajectory inference is maturing, primarily for linear and bifurcating trajectories. Nonetheless, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable and fast tool for analysing single-cell -omics datasets with complex trajectories. Foremost, new methods should focus on improving the inference of the topology for tree and more complex topologies, and should include better ordering methods, perhaps borrowed from the top linear TI methods. Furthermore, new tools should incorporate higher standards for code assurance and documentation, to improve the friendliness towards both users and developers. To support the development of these new tools, we provide all datasets from this study [@cannoodt_goldstandardsinglecell_] and provide the code to compare methods in a series of R packages available at [github.com/dynverse/dynverse](https://github.com/dynverse/dynverse).

@@

* **Random:** A random network with 50 milestones according to the Barabási–Albert model is generated, and cells are placed randomly along the network. This method provides a baseline performance to compare against; each other TI method should be able to obtain a higher benchmark score than this control method.
* **Component 1:**: This method returns the first component of a PCA dimensionality reduction as a linear trajectory. We added this method because indeed in some studies the first component of a dimensionality reduction is being used to pseudotemporally order cells [@kouno_temporaldynamicstranscriptional_2013,@zeng_pseudotemporalorderingsingle_2017].
* **Angle:** Similar to the previous method, this method computes the angle with respect to the origin in a two-dimensional PCA plot, and returns a cyclical trajectory. As reCAT was the only method specifically designed to produce cyclical trajectories, we believed it would be important to add this simple approach.

## Real datasets
We gathered `r sum(tasks_info$task_source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`). The scripts to download and process these datasets are available on our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalised and filtered using a common pipeline, discussed later. We determined a reference standard for every dataset using labelling provided by the author's, and classified the standards into gold and silver based on whether this labelling was determined by the expert using the expression (silver standard) or using other external information (such as FACS or the origin of the sample, gold standard) (`r ref("stable", "datasets")`).

@@

```{r}
add_sfig(
  figure_file("edge_flip_comparison.svg", experiment_id="3-metric_characterisation/4-edge_flip"), 
  "edge_flip_comparison", 
  paste0("Edge flip score values for common trajectory topologies"), 
  15, 
  5
)
```

@@

We created a transparent scoring scheme to check the quality of each method based on several existing tool quality and programming guidelines in literature and online (`r ref("stable", "qc_checks")`). The goal of this quality control in the first place is to stimulate the improvement of current methods, and the development of user and developer friendly new methods. The quality control assessed 6 categories, each looking at several aspects, which are further divided into individual items. The availability category checks whether the method is easily available, whether the code and dependencies can be easily installed, and how the method can be used. The code quality assesses the quality of the code both from a user perspective (function naming, dummy proofing and availability of plotting functions) and a developer perspective (consistent style and code duplication). The code assurance category is frequently overlooked, and checks for code testing, continuous integration [@beaulieu-jones_reproducibility_2017] and an active support system. The documentation category checks the quality of the documentation, both externally (tutorials and function documentation) and internally (inline documentation). The behaviour category assesses the ease by which the method can be run, by looking for unexpected output files and messages, prior information and how easy the trajectory model can be extracted from the output. Finally, we also assessed certain aspects of the study in which the method was proposed, such as publication in a peer-reviewed journal, the number of dataset in which the usefulness of the method was shown, and the scope of method evaluation in the paper.

@@

## Trajectory types
We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria (`r ref("fig", "ti_evaluation_overview", "c")`). These trajectory types start from the most general trajectory type, a disconnected directed graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cyclical. A disconnected directed graph is a graph in which only one edge can exist between two nodes. A directed graph is a disconnected graph in which all nodes are (indirectly) connected. A directed acyclic graph is a graph containing no cycles. A rooted tree is a directed acyclic graph containing no convergences (no nodes with in-degree higher than 1). A rooted binary tree is a rooted tree with every node having maximally two outgoing edges. A convergence is a directed acyclic graph in which only one node has a degree larger than one and this same node has an indegree of one. A multifurcation is a rooted tree in which only one node has a degree larger than one. A bifurcation is a multifurcation in which only one node has a degree equal to 3. A directed linear graph is a graph in which no node has a degree larger than 3. Finally, a directed cycle is a directed graph in which no node has a degree larger than 3 but contains one cycle. For every directed trajectory type, a corresponding undirected trajectory type can also be defined, with directed acyclic graphs and directed graphs merging to undirected graph, and convergence and bifurcation merging to a simple fork. In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with the only exception being DPT (limited to bifurcating trajectories).

## Data and code availability
The datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533)). The code to fully reproduce the results from this study has been split into several R packages and is available on github ([github.com/dynverse/dynverse](https://github.com/dynverse/dynverse)).

```{r "supplementary", echo=FALSE, results='asis'}
if (params$table_format == "html") 
  source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
write_rds(lst(refs, sfigs, stables), derived_file("sfigs_stables.rds", "paper"))
```
